Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-28 analysis/paper/paper.Rmd
add references

Pottery analyzed in this paper comes from Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between. This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century. The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China.

The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site.

>>>>

Pottery analyzed in this paper comes from Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007]. This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century. The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
